---
title: 'Part 1: A Simulation Exercise'
author: "Yasser Gonzalez -- http://yassergonzalez.com"
date: "February 2015"
output:
  pdf_document: default
---

# Overview

In the first part of the project we study the exponential distribution in R 
and compare it with the Central Limit Theorem. The probability density function
of the exponential distribution is given by

\begin{equation*}
f(x;\lambda) = \begin{cases}
\lambda e^{-\lambda x} & x \ge 0, \\
0 & x < 0.
\end{cases},
\end{equation*}

where $\lambda > 0$ is the rate parameter. The distribution mean is $1/\lambda$ 
and the standard deviation is also $1/\lambda$. 

This report investigates the distribution of 1,000 observations of averages of 
40 simulations of the exponential distribution. The rate parameter $\lambda$ will 
be set to 0.2 for all of the simulations.

# Simulations

The first step before running the simulations is to set the random seed to a fixed 
number to make the results reproducible. Also, the R packages that will be used 
throughout the report are loaded into the interpreter.

```{r}
set.seed(12345)

library("ggplot2")
```

We must obtain a sample with 1,000 observations of averages of 40 simulations 
of the exponential distribution, which can be generated in R with the following 
code fragment.

```{r}
exp_simulations <- replicate(1000, mean(rexp(40, rate = 0.2)))
```

After running this code, `exp_simulations` contains a vector with 1,000 elements.
This vector constitutes the sample that will be studied in the remainder 
of the report.

# Sample Mean versus Theoretical Mean

The expected value of the sample mean of the averages of the simulations of the 
exponential distribution is the theoretical distribution mean---i.e. $1/\lambda$, 
in this particular case $1/0.2 = 5$. Therefore, if we calculate the sample mean of
`exp_simulations` we must obtain a value close to 5.

```{r}
mean(exp_simulations)
```

This is also illustrated in the following plot, showing a histogram of 
`exp_simulations` together with vertical lines for the sample mean (blue)
and the theoretical mean (red).

```{r sample_vs_theoretical_mean}
df <- data.frame(exp_simulations = exp_simulations)
ggplot(df, aes(x = exp_simulations)) +
    geom_histogram(binwidth = 0.2) +
    geom_vline(xintercept = mean(exp_simulations), colour = "blue") +
    geom_vline(xintercept = 1 / 0.2, colour = "red") +
    ggtitle("Sample Mean versus Theoretical Mean") +
    xlab("Average of 40 simulations of the exponential distribution") +
    ylab("Count")
```

# Sample Variance versus Theoretical Variance

# Distribution
