---
title: 'Part 1: A Simulation Exercise'
author: "Yasser Gonzalez -- http://yassergonzalez.com"
date: "February 2015"
output:
  pdf_document: default
---

# Overview

In the first part of the project we study the exponential distribution in R 
and compare it with the Central Limit Theorem. The probability density function
of the exponential distribution is given by

\begin{equation*}
f(x;\lambda) = \begin{cases}
\lambda e^{-\lambda x} & x \ge 0, \\
0 & x < 0.
\end{cases},
\end{equation*}

where $\lambda > 0$ is the rate parameter. The distribution mean is $1/\lambda$ 
and the standard deviation is also $1/\lambda$. 

This report investigates the distribution of 1,000 observations of averages of 
40 simulations of the exponential distribution. The rate parameter $\lambda$ will 
be set to 0.2 for all of the simulations.

# Simulations

The first step before running the simulations is to set the random seed to a fixed 
number to make the results reproducible. Also, the R packages that will be used 
throughout the report are loaded into the interpreter.

```{r}
set.seed(12345)

library("ggplot2")
```

We must obtain a sample with 1,000 observations of the mean of 40 simulations 
of the exponential distribution, which can be generated in R with the following 
code fragment.

```{r}
exp_simulations <- replicate(1000, mean(rexp(40, rate = 0.2)))
```

After running this code, `exp_simulations` contains a vector with 1,000 elements.
This vector constitutes the population that will be studied in the remainder 
of the report.

# Sample Mean versus Theoretical Mean

# Sample Variance versus Theoretical Variance

# Distribution
